---
title: "Project3"
subtitle: "Ship tracking from bearings"
author: "Florian Beiser, Yaolin Ge"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(MASS)
```

# Introduction

The state of the problem is $\mathrm{x}_t = (E_t, N_t, v_t, u_t)$, where $x_1 \sim \mathcal{N}_4(\mu_1,\Sigma_1)$.

The forward model is specified by $\mathrm{x}_{t+1} = \mathbf{A} \mathrm{x}_t + \varepsilon_{t+1}$ for $t=1,\dots,T-1$. 

The observation model is specified by $\mathrm{y}_t = H(\mathrm{x}_t) + \beta_t$ for $t=1,\dots,T$ with $H(\mathrm{x}_t) = (\arctan(E_t/N_t),\arctan((40-N_t)/(40-E_t)))^\top$

```{r intro}
set.seed(202104)

# Prior
mu1 = c(10,30,10,-10)
Sigma1 = diag(c(10**2,10**2,5**2,5**2))

# Forward model 
delta = 1/60

A = diag(4)
A[1,3] = delta 
A[2,4] = delta 

Eps_model = diag(c(0.1**2,0.1**2,0.5**2,0.5**2))

ModelStep <- function(x){
  x = A%*%x #+ mvrnorm(mu=rep(0,4), Sigma=Eps_model)
  return(x)
}


# Observation model
Eps_obs = diag(c(0.1**2,0.1**2))

Observation <- function(x){
  y = c(atan(x[1]/x[2]), atan((40-x[2])/(40-x[1]))) #+ mvrnorm(mu=rep(0,2), Sigma=Eps_obs)
  return(y)
}


# Observations 
obsA = read.table("sensorA.txt")
obsB = read.table("sensorB.txt")

obs = cbind(obsA,obsB)

## Visualisation
par(mfrow=c(1,2))
plot(obs[,1], main="Observations from sensor A",
     xlab="time", ylab="angle")
plot(obs[,2], main="Observations from sensor B",
     xlab="time", ylab="angle")

```



# Extended Kalman filter

For the extended Kalman filter, we linearize the observation operator $H$ and construct its Jacobian $\mathbf{H}|_x$.

```{r EKF}

linObsOp <- function(x){
  H = matrix(0, ncol=4, nrow=2)
  H[1,1] = x[2]/(x[1]**2 + x[2]**2)
  H[1,2] = -x[1]/(x[2]**2 + x[1]**2)
  H[2,1] = (40-x[2])/(x[1]**2 - 80*x[1] + x[2]**2 - 80*x[2] + 3200)
  H[2,2] = (x[2]-40)/(x[1]**2 - 80*x[1] + x[2]**2 - 80*x[2] + 3200)
  return(H)
}

# Bookkeeping 
T = 50
mus    = array(0, dim=c(4,T))
Sigmas = array(0, dim=c(4,4,T))

# DATA ASSIMILATION in time loop
mu = mu1
Sigma = Sigma1
for (t in 1:T){
  ## Forward Step
  # NOTE: the initial prior is already given for the first observation time
  if (t != 1){
    mu = ModelStep(mu)
  }
  Sigma = A %*% Sigma %*% t(A) + Eps_model
  
  ## Analysis Step
  H = linObsOp(mu)
  S = H %*% Sigma %*% t(H) + Eps_obs
  K = Sigma %*% t(H) %*% solve(S)
  
  # Updating
  y = t(obs[t,]) 
  mu = mu + K %*% (y - Observation(mu))
  Sigma = (diag(4) - K %*% H) %*% Sigma 
  
  # Storage 
  mus[,t] = mu
  Sigmas[,,t] = Sigma
}


plot(mus[1,],mus[2,], main="Filtered position means",
     xlab="East", ylab="North",
     xlim=c(0,40), ylim=c(0,40),
     col="red", type="l")

lines(mus[1,] - 2*sqrt(Sigmas[1,1,]), mus[2,] - 2*sqrt(Sigmas[2,2,]))
lines(mus[1,] + 2*sqrt(Sigmas[1,1,]), mus[2,] + 2*sqrt(Sigmas[2,2,]))

par(mfrow=c(1,2))
plot(Sigmas[1,1,], main="Filtered position variances (east)",
     xlab="time", ylab="variance")
plot(Sigmas[2,2,], main="Filtered position variances (north)",
     xlab="time", ylab="variance")

plot(mus[3,], main="Filtered velocity means (east)",
     xlab="time", ylab="velocity")
plot(mus[4,], main="Filtered velocity means (east)",
     xlab="time", ylab="velocity")
```


# Ensemble Kalman Filter

The ensemble version replaces the covariance matrix by an estimate of it.

```{r EnKF}

# Bookkeeping 
B = 1000
mus    = array(0, dim=c(4,T))
Sigmas = array(0, dim=c(4,4,T))


# DATA ASSIMILATION in time loop
mu = mu1
Sigma = Sigma1

for (t in 1:T){
  ## Forward Step
  # NOTE: the initial prior is already given for the first observation time
  if (t == 1){
    ensemble = t(mvrnorm(B, mu=mu, Sigma=Sigma))
  }
  else{
    for (e in 1:B){
      ensemble[,e] = ModelStep(ensemble[,e]) + mvrnorm(mu=rep(0,4), Sigma=Eps_model)
    }
  }
  ensemble_mean = rowMeans(ensemble)
  Sigma = 1/(B-1)*(ensemble-ensemble_mean)%*%t(ensemble-ensemble_mean)
  
  ## Analysis Step
  for (e in 1:B){
    # Analysing single ensemble member
    H = linObsOp(ensemble[,e])
    S = H %*% Sigma %*% t(H) + Eps_obs
    K = Sigma %*% t(H) %*% solve(S)
    
    # Updating single ensemble member
    y = t(obs[t,]) + mvrnorm(mu=rep(0,2), Sigma=Eps_obs)
    d = y - Observation(ensemble[,e])
    ensemble[,e] = ensemble[,e] + K %*% d
  }
  
  # Storage 
  mus[,t] = rowMeans(ensemble)
  Sigmas[,,t] = Sigma
}

plot(mus[1,],mus[2,], main="Filtered position means",
     xlab="East", ylab="North",
     xlim=c(0,40), ylim=c(0,40))

par(mfrow=c(1,2))
plot(Sigmas[1,1,], main="Filter position variances (east)",
     xlab="time", ylab="variance")
plot(Sigmas[2,2,], main="Filter position variances (north)",
     xlab="time", ylab="variance")

plot(mus[3,], main="Filter velocity means (east)",
     xlab="time", ylab="velocity")
plot(mus[4,], main="Filter velocity means (east)",
     xlab="time", ylab="velocity")
```